---
title: Interacting with the gypsum REST API
author:
- name: Aaron Lun
  email: infinite.monkeys.with.keyboards@gmail.com
package: gypsum
date: "Revised: December 5, 2023"
output:
  BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{Hitting the gypsum API}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo=FALSE}
library(BiocStyle)
self <- Githubpkg("ArtifactDB/gypsum");
knitr::opts_chunk$set(error=FALSE, warning=FALSE, message=FALSE)
```

# Introduction

The `r self` package implements an R client for the [API of the same name](https://gypsum-test.aaron-lun.workers.dev).
This allows Bioconductor packages to easily store and retrieve data from the **gypsum** backend.
It also provides mechanisms to allow package maintainers to easily manage upload authorizations and third-party contributions.
Readers are referred to [API's documentation](https://github.com/ArtifactDB/gypsum-worker) for a description of the concepts;
this guide will strictly focus on the usage of the `r self` package.

# Reading files

`r self` provides several convenience methods for reading from the **gypsum** bucket:

```{r}
library(gypsum)
listAssets("test-R")
listVersions("test-R", "basic")
listFiles("test-R", "basic", "v1")

out <- saveFile("test-R", "basic", "v1", "blah.txt")
readLines(out)

dir <- saveVersion("test-R", "basic", "v1")
list.files(dir, all.files=TRUE, recursive=TRUE)
```

We can also fetch the summaries and manifests for each version of a project's assets.

```{r}
fetchManifest("test-R", "basic", "v1")
fetchSummary("test-R", "basic", "v1")
```

All read operations involve a publicly accessible bucket so no authentication is required.

# Uploading files

To demonstrate, let's say we have a directory of files that we wish to upload to the backend.

```{r}
tmp <- tempfile()
dir.create(tmp)
write(file=file.path(tmp, "foo"), letters)
write(file=file.path(tmp, "bar"), LETTERS)
write(file=file.path(tmp, "whee"), 1:10)
```

We run the upload sequence of `startUpload()`, `uploadFiles()` and `completeUpload()`.
This requires authentication via GitHub, which is usually prompted but can also be set beforehand via `setAccessToken()` (e.g., for batch jobs).

```r
init <- startUpload(
    project=project_name,
    asset=asset_name,
    version=version_name,
    files=list.files(tmp, recursive=TRUE),
    directory=tmp
)

tryCatch({
    uploadFiles(init, directory=tmp)
    completeUpload(init)
}, error=function(e) {
    abortUpload(init) # clean up if the upload fails.
    stop(e)
})
```

More advanced developers may consider using `links=` in `startUpload()` to improve efficiency by deduplicating redundant files on the **gypsum** backend.
We can also set `concurrent=` to parallelize the uploads in `uploadFiles()`.

# Changing permissions

Upload authorization is determined by each project's permissions, which are controlled by project owners.
Both uploaders and owners are identified based on their GitHub logins:

```{r}
fetchPermissions("test-R")
```

Owners can add more uploaders (or owners) via the `setPermissions()` function.
Uploaders can be scoped to individual assets or versions, and an expiry date may be attached to each authorization:

```r
setPermissions("test-R", uploaders=list(
    list(
        id="jkanche", 
        until=Sys.time() + 24 * 60 * 60,
        asset="jays-happy-fun-time",
        version="1"
    )
))
```

Organizations may also be added in the permissions, in which case the authorization extends to all members of that organization.

# Probational uploads

Unless specified otherwise, all uploaders are considered to be "untrusted".
Any uploads from such untrusted users are considered "probational" and must be approved by the project owners before they are considered complete.
Alternatively, an owner may reject an upload, which deletes all the uploaded files from the backend.

```r
approveProbation("test-R", "third-party-upload", "good")
rejectProbation("test-R", "third-party-upload", "bad")
```

An uploader can be trusted by setting `trusted=TRUE` in `setPermissions()`.
Owners and trusted uploaders may still perform probational uploads (e.g., for testing) by setting `probation=TRUE` in `startUpload()`.

# Validating metadata

Databases can operate downstream of the **gypsum** backend to create performant search indices, usually based on special metadata files.
`r self` provides some utilities to check that metadata follows the JSON schema of some known downstream databases.

```{r}
# Fetching for the gypsum-Bioconductor database
schema <- fetchMetadataSchema("bioconductor")
cat(head(readLines(schema)), sep="\n")
```

Uploaders can verify that their metadata respects this schema via the `validateMetadata()` function.
This ensures that the uploaded files can be successfully indexed by the database, given that the **gypsum** backend itself applies no such checks.

```{r}
metadata <- list(                           
    title="Fatherhood",
    description="Luke ich bin dein Vater.",
    sources=list(
       list(provider="GEO", id="GSE12345")
    ),
    taxonomy_id=list("9606"),
    genome=list("GRCm38"),
    maintainer_name="Darth Vader",
    maintainer_email="vader@empire.gov",
    bioconductor_version="3.10"
)

validateMetadata(metadata, schema)
```

# Administration

Administrators of the **gypsum** instance can create projects with new permissions:

```r
createProject("my-new-project", 
    owners="jkanche", 
    uploaders=list(
        list(
            id="LTLA", 
            asset="aarons-stuff"
        )
    )
)
```

Administrators may also delete projects, assets or versions, though this should be done sparingly as it violates **gypsum**'s expectations of immutability.

```r
removeProject("my-new_project")
```

# Session information {-}

```{r}
sessionInfo()
```

